---
title: "R Basics Tutorial - Data Analysis for Criminology"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Getting Started with R

## 1.1 Creating and Working with Text Files

### Create a simple text file
```{r create-text-file}
# Create a text file with some content
text_content <- c("This is line 1",
                  "This is line 2",
                  "This is line 3")

# Write to a text file
writeLines(text_content, "sample_text.txt")

# Read the text file
read_content <- readLines("sample_text.txt")
print(read_content)
```

### Create a CSV file manually
```{r create-csv}
# Create sample criminology data
crime_data <- data.frame(
  ID = 1:5,
  Age = c(25, 32, 19, 45, 28),
  Gender = c("Male", "Female", "Male", "Male", "Female"),
  Crime_Type = c("Theft", "Assault", "Vandalism", "Fraud", "Theft"),
  Sentence_Months = c(12, 24, 6, 36, 18)
)

# Save as CSV
write.csv(crime_data, "crime_data.csv", row.names = FALSE)
```

---

# 2. Loading and Reading Data

## 2.1 Loading CSV Files

### Method 1: Using base R
```{r load-csv-base}
# Read CSV file
data_base <- read.csv("crime_data.csv")
print(data_base)
```

### Method 2: Using readr package (better for large files)
```{r load-csv-readr}
# Install readr if needed (uncomment below)
# install.packages("readr")

library(readr)
data_readr <- read_csv("crime_data.csv")
print(data_readr)
```

### Method 3: Using file.choose() for interactive selection
```{r load-csv-interactive, eval=FALSE}
# This opens a file browser window
my_data <- read.csv(file.choose())
```

---

# 3. Understanding and Using Libraries (Packages)

## 3.1 What is a Library?

A library (or package) is a collection of pre-written functions and datasets that extend R's capabilities.

## 3.2 Installing Packages

```{r install-packages, eval=FALSE}
# Install a single package
install.packages("dplyr")

# Install multiple packages
install.packages(c("ggplot2", "tidyverse", "haven"))

# Install from CRAN
install.packages("psych")
```

## 3.3 Loading Packages

```{r load-packages}
# Load a package for use in your session
library(dplyr)
library(ggplot2)

# Alternative way (less common)
require(tidyverse)
```

## 3.4 Checking Installed Packages

```{r check-packages, eval=FALSE}
# See all installed packages
installed.packages()

# Check if a specific package is installed
"dplyr" %in% rownames(installed.packages())
```

---

# 4. Examining Datasets

## 4.1 Using view() Function

```{r view-data, eval=FALSE}
# Open data in a spreadsheet-like viewer (RStudio only)
View(crime_data)

# Alternative lowercase version
view(crime_data)
```

## 4.2 Using str() - Structure of Data

```{r str-function}
# Show structure: variable types, number of observations
str(crime_data)
```

**Output explains:**
- `data.frame`: Type of object
- `5 obs. of 5 variables`: 5 rows and 5 columns
- Variable types: `int` (integer), `chr` (character), `num` (numeric)

## 4.3 Using names() - Column Names

```{r names-function}
# Get column names
names(crime_data)

# Alternative
colnames(crime_data)
```

## 4.4 Using dim() - Dimensions

```{r dim-function}
# Get dimensions (rows, columns)
dim(crime_data)

# Number of rows
nrow(crime_data)

# Number of columns
ncol(crime_data)
```

## 4.5 Using glimpse() - Quick Overview

```{r glimpse-function}
library(dplyr)

# Transposed view of data
glimpse(crime_data)
```

## 4.6 Other Useful Functions for Examining Data

```{r other-examine}
# First 6 rows
head(crime_data)

# Last 6 rows
tail(crime_data)

# First n rows
head(crime_data, n = 3)

# Summary statistics
summary(crime_data)
```

---

# 5. Working with Missing Values

## 5.1 Creating Data with Missing Values

```{r create-missing}
# Create data with missing values (NA)
crime_data_missing <- data.frame(
  ID = 1:6,
  Age = c(25, NA, 19, 45, 28, NA),
  Gender = c("Male", "Female", NA, "Male", "Female", "Male"),
  Sentence_Months = c(12, 24, NA, 36, 18, 6)
)

print(crime_data_missing)
```

## 5.2 Detecting Missing Values

```{r detect-missing}
# Check for NA values
is.na(crime_data_missing)

# Count NA values per column
colSums(is.na(crime_data_missing))

# Total number of NA values
sum(is.na(crime_data_missing))

# Which rows have NA values?
complete.cases(crime_data_missing)

# Show only complete cases
crime_data_missing[complete.cases(crime_data_missing), ]
```

## 5.3 Handling Missing Values

```{r handle-missing}
# Remove rows with any NA values
clean_data <- na.omit(crime_data_missing)
print(clean_data)

# Replace NA with a specific value
crime_data_missing$Age[is.na(crime_data_missing$Age)] <- mean(crime_data_missing$Age, na.rm = TRUE)

# Using tidyverse to drop NA
library(tidyr)
crime_data_clean <- crime_data_missing %>% drop_na()
```

---

# 6. Coding and Recoding Variables

## 6.1 Creating New Variables

```{r create-variables}
# Create a new variable based on existing ones
crime_data$Age_Group <- ifelse(crime_data$Age < 30, "Young", "Adult")
print(crime_data)

# Create categorical variable
crime_data$Severity <- ifelse(crime_data$Sentence_Months > 20, "High", "Low")
print(crime_data)
```

## 6.2 Recoding Variables - Multiple Categories

```{r recode-multiple}
# Using nested ifelse
crime_data$Age_Category <- ifelse(crime_data$Age < 25, "Youth",
                                   ifelse(crime_data$Age < 35, "Young Adult",
                                          "Adult"))

# Using cut() for numeric to categorical
crime_data$Age_Bins <- cut(crime_data$Age, 
                            breaks = c(0, 25, 35, 100),
                            labels = c("Youth", "Young Adult", "Adult"))

print(crime_data)
```

## 6.3 Recoding with dplyr

```{r recode-dplyr}
library(dplyr)

crime_data <- crime_data %>%
  mutate(
    # Recode using case_when (more flexible)
    Crime_Severity = case_when(
      Crime_Type %in% c("Theft", "Vandalism") ~ "Minor",
      Crime_Type == "Assault" ~ "Moderate",
      Crime_Type == "Fraud" ~ "Serious",
      TRUE ~ "Other"
    )
  )

print(crime_data)
```

## 6.4 Recoding Factor Levels

```{r recode-factors}
# Original factor
crime_data$Gender <- factor(crime_data$Gender)
levels(crime_data$Gender)

# Recode factor levels
crime_data$Gender_Coded <- recode(crime_data$Gender, 
                                   "Male" = "M", 
                                   "Female" = "F")
print(crime_data)
```

---

# 7. Creating Vectors

## 7.1 Numeric Vectors

```{r create-numeric-vectors}
# Create numeric vector
ages <- c(25, 32, 19, 45, 28)
print(ages)

# Sequence
id_sequence <- 1:10
print(id_sequence)

# Using seq()
custom_seq <- seq(from = 0, to = 100, by = 10)
print(custom_seq)

# Repeat values
repeated <- rep(5, times = 6)
print(repeated)
```

## 7.2 Character Vectors

```{r create-character-vectors}
# Create character vector
names_vector <- c("John", "Jane", "Bob", "Alice", "Eve")
print(names_vector)

# Cities
cities <- c("Brussels", "Ghent", "Antwerp", "Bruges")
print(cities)
```

## 7.3 Logical Vectors

```{r create-logical-vectors}
# Logical vector
convicted <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
print(convicted)

# Created from conditions
high_risk <- ages > 30
print(high_risk)
```

---

# 8. Working with Factors

## 8.1 What is a Factor?

Factors are used to represent categorical data with fixed levels.

## 8.2 Creating Factors

```{r create-factors}
# Create a factor from character vector
gender <- c("Male", "Female", "Male", "Male", "Female")
gender_factor <- factor(gender)
print(gender_factor)

# Check levels
levels(gender_factor)

# Create ordered factor
education <- c("High School", "Bachelor", "Master", "High School", "PhD")
education_factor <- factor(education, 
                           levels = c("High School", "Bachelor", "Master", "PhD"),
                           ordered = TRUE)
print(education_factor)
```

## 8.3 Converting Between Types

```{r convert-factors}
# Character to factor
crime_types <- c("Theft", "Assault", "Fraud", "Theft")
crime_factor <- as.factor(crime_types)

# Factor to character
crime_char <- as.character(crime_factor)

# Factor to numeric (be careful!)
levels_numeric <- as.numeric(crime_factor)
print(levels_numeric)  # Returns level numbers, not original values
```

## 8.4 Modifying Factor Levels

```{r modify-factors}
# Rename levels
crime_factor_renamed <- factor(crime_types)
levels(crime_factor_renamed) <- c("Physical Violence", "Financial Crime", "Property Crime")
print(crime_factor_renamed)

# Add new levels
crime_factor_expanded <- factor(crime_types, levels = c("Theft", "Assault", "Fraud", "Vandalism"))
print(crime_factor_expanded)
```

---

# 9. Creating Frequency and Percentage Tables

## 9.1 Frequency Tables

```{r frequency-tables}
# Simple frequency table
freq_crime <- table(crime_data$Crime_Type)
print(freq_crime)

# Two-way frequency table (cross-tabulation)
freq_cross <- table(crime_data$Gender, crime_data$Crime_Type)
print(freq_cross)
```

## 9.2 Percentage Tables

```{r percentage-tables}
# Percentage table
percent_crime <- prop.table(table(crime_data$Crime_Type)) * 100
print(percent_crime)

# Round percentages
percent_crime_rounded <- round(percent_crime, 2)
print(percent_crime_rounded)
```

## 9.3 Combined Frequency and Percentage Tables

```{r combined-tables}
# Create a combined table
crime_freq <- table(crime_data$Crime_Type)
crime_prop <- prop.table(crime_freq) * 100

combined_table <- data.frame(
  Frequency = as.vector(crime_freq),
  Percentage = round(as.vector(crime_prop), 2)
)
rownames(combined_table) <- names(crime_freq)
print(combined_table)
```

## 9.4 Using dplyr for Tables

```{r dplyr-tables}
library(dplyr)

# Frequency and percentage with dplyr
crime_summary <- crime_data %>%
  group_by(Crime_Type) %>%
  summarise(
    Frequency = n(),
    Percentage = round(n() / nrow(crime_data) * 100, 2)
  ) %>%
  arrange(desc(Frequency))

print(crime_summary)
```

## 9.5 Advanced Tables with janitor Package

```{r janitor-tables}
# Install janitor if needed
# install.packages("janitor")

library(janitor)

# Create a nice frequency table
crime_data %>%
  tabyl(Crime_Type) %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_totals("row")

# Cross-tabulation with percentages
crime_data %>%
  tabyl(Gender, Crime_Type) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()
```

---

# 10. Practical Example: Complete Workflow

```{r complete-example}
# 1. Create a larger criminology dataset
set.seed(123)  # For reproducibility

offender_data <- data.frame(
  ID = 1:50,
  Age = sample(18:65, 50, replace = TRUE),
  Gender = sample(c("Male", "Female"), 50, replace = TRUE),
  Education = sample(c("Elementary", "High School", "Bachelor", "Master"), 
                     50, replace = TRUE, prob = c(0.1, 0.5, 0.3, 0.1)),
  Prior_Convictions = sample(0:5, 50, replace = TRUE),
  Crime_Type = sample(c("Theft", "Assault", "Fraud", "Vandalism", "Drug Offense"), 
                      50, replace = TRUE),
  Sentence_Days = sample(30:1095, 50, replace = TRUE)
)

# 2. Save the dataset
write.csv(offender_data, "offender_data.csv", row.names = FALSE)

# 3. Load the dataset
offender_data <- read.csv("offender_data.csv")

# 4. Examine the data
str(offender_data)
summary(offender_data)
head(offender_data)

# 5. Create new variables
offender_data <- offender_data %>%
  mutate(
    Age_Group = cut(Age, breaks = c(0, 25, 40, 100), 
                    labels = c("Young", "Middle-Aged", "Older")),
    High_Risk = ifelse(Prior_Convictions > 2, "Yes", "No"),
    Sentence_Category = case_when(
      Sentence_Days < 180 ~ "Short",
      Sentence_Days < 730 ~ "Medium",
      TRUE ~ "Long"
    )
  )

# 6. Create frequency tables
print("Crime Type Frequency:")
table(offender_data$Crime_Type)

print("\nAge Group by Crime Type:")
table(offender_data$Age_Group, offender_data$Crime_Type)

# 7. Create summary table
offender_summary <- offender_data %>%
  group_by(Crime_Type, Gender) %>%
  summarise(
    Count = n(),
    Avg_Age = round(mean(Age), 1),
    Avg_Sentence = round(mean(Sentence_Days), 0),
    .groups = "drop"
  )

print(offender_summary)
```

---

# 11. Quick Reference Cheat Sheet

## Common Functions

| Function | Purpose | Example |
|----------|---------|---------|
| `read.csv()` | Load CSV file | `data <- read.csv("file.csv")` |
| `write.csv()` | Save CSV file | `write.csv(data, "file.csv")` |
| `View()` | Open data viewer | `View(data)` |
| `str()` | Show structure | `str(data)` |
| `names()` | Column names | `names(data)` |
| `dim()` | Dimensions | `dim(data)` |
| `head()` | First rows | `head(data)` |
| `summary()` | Summary stats | `summary(data)` |
| `table()` | Frequency table | `table(data$column)` |
| `is.na()` | Check missing | `is.na(data)` |
| `na.omit()` | Remove missing | `na.omit(data)` |
| `factor()` | Create factor | `factor(vector)` |
| `c()` | Create vector | `c(1, 2, 3)` |
| `ifelse()` | Conditional | `ifelse(condition, yes, no)` |

## Useful Packages

- **tidyverse**: Data manipulation and visualization
- **dplyr**: Data manipulation
- **ggplot2**: Data visualization
- **haven**: Import SPSS/Stata files
- **readr**: Fast CSV reading
- **janitor**: Clean data and tables
- **psych**: Psychological/statistical functions

---

# 12. Exercises

Try these exercises to practice:

1. Create a CSV file with 10 fictional offenders including age, gender, and crime type
2. Load the file and examine it using `str()`, `summary()`, and `glimpse()`
3. Create a new variable categorizing ages into "Young" (<30) and "Adult" (≥30)
4. Convert gender to a factor
5. Create a frequency table showing crime types
6. Calculate the percentage of each crime type
7. Create a cross-tabulation of gender and crime type

---

**End of Tutorial**
