---
title: "R for Criminology Students: Complete Course Plan"
author: "Course Guide"
date: "`r Sys.Date()`"
output: 
  word_document:
    toc: true
    toc_depth: 3
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Introduction

This document outlines a **10-session track** (±2h each) plus a final project for teaching R to criminology bachelors with zero coding experience. The goal is to move **from installation → analysis → APA tables → full thesis/paper in RMarkdown with Word/PDF output**.

---

# Phase 0 – Installation & Setup (before first class)

**Goal:** Every student has a working R/RStudio setup and can open RStudio.

## 0.1. Software

Give students a 1-page PDF with screenshots:

1. **Install R**
   * Go to CRAN (cloud.r-project.org)
   * Download R for Windows/Mac and install with defaults.

2. **Install RStudio Desktop**
   * Download RStudio Desktop (free) and install.

3. **(Optional for PDF output)** Install tinytex
   * Later, you'll run in R:

```{r install-tinytex}
install.packages("tinytex")
tinytex::install_tinytex()
```

4. **Create one folder for the course**
   * E.g. `Documents/R_criminology_course/`
   * All datasets + R scripts + Rmd files go here.

## 0.2. First startup checklist in class

In the **first 15 minutes of session 1**, have them verify:

* RStudio opens.
* They can create:
  * A new **R Script** (`File → New File → R Script`).
  * A new **R Markdown** document (`File → New File → R Markdown…`).

---

# Phase 1 – Basic R & RStudio (Sessions 1–2)

## Session 1 – R as a tool for criminology

### Goals

* Lower anxiety, show use cases.
* Basic syntax: objects, functions, vectors.

### Core content

* Very short intro: *Why R for criminology?* (reproducibility, transparency, free).
* RStudio tour: Console, Script, Environment, Plots, Files.
* R as a calculator:

```{r calculator}
2 + 2
10 / 2
3^2
```

* Objects:

```{r objects} git
x <- 10
y <- 5
x + y
```

* Vectors (use crime-related examples):

```{r vectors}
burglary_counts <- c(12, 8, 15, 9, 11)
mean(burglary_counts)
sd(burglary_counts)
```

* Commenting:

```{r commenting}
# This is the number of burglaries in 5 neighbourhoods
```

### Product

Saved script `session1_intro.R` with comments + simple calculations.

---

## Session 2 – Data frames & criminology datasets

### Goals

* Understand data frames (rows = cases, columns = variables).
* Load and inspect a CSV.

### Core content

* Introduce packages:

```{r tidyverse-install}
install.packages("tidyverse")
library(tidyverse)
```

* Import survey data (you provide something like `fear_of_crime.csv`):

```{r import-data}
survey <- read_csv("fear_of_crime.csv")
```

* Inspect:

```{r inspect-data}
View(survey)
names(survey)
dim(survey)
head(survey)
summary(survey)
str(survey)
```

* Access variables:

```{r access-variables}
survey$fear
mean(survey$fear, na.rm = TRUE)
```

### Homework idea

* Students answer simple questions with R:
  * How many respondents?
  * What is the mean fear?
  * What variables are in the dataset?

---

# Phase 2 – Data Wrangling & Descriptives (Sessions 3–4)

## Session 3 – Selecting & filtering data

### Goals

* Teach them to subset and create variables with `dplyr`.

### Core content

* `select()`, `filter()`, `mutate()`, logical operators.
* Example:

```{r dplyr-basics}
library(dplyr)

small <- survey %>%
  select(gender, age, fear, victim, trust_police)

victims <- small %>%
  filter(victim == 1)

small <- small %>%
  mutate(
    high_fear = ifelse(fear >= 4, 1, 0),
    older_30  = ifelse(age > 30, 1, 0)
  )
```

### Homework idea

* Create a new variable (e.g. `low_trust`) and filter a subgroup (e.g. victims with low trust).
* Answer 2–3 short questions about this subgroup.

---

## Session 4 – Descriptive statistics & simple tables

### Goals

* Move from "data" to "results" (summary stats).

### Core content

* Frequency tables:

```{r freq-tables}
table(small$gender)
prop.table(table(small$gender))
```

* With `dplyr`:

```{r dplyr-freq}
small %>% count(gender) %>%
  mutate(prop = n / sum(n))
```

* Grouped means:

```{r grouped-means}
small %>%
  group_by(gender) %>%
  summarise(
    mean_fear = mean(fear, na.rm = TRUE),
    sd_fear   = sd(fear, na.rm = TRUE),
    n         = n()
  )
```

* Interpret results in simple criminological language.

### Homework

* 3–4 questions like:
  * Is mean fear higher among victims?
  * Which neighbourhood has highest fear?

---

# Phase 3 – Inferential Analysis (Sessions 5–6)

## Session 5 – T-tests & chi-square with R

### Goals

* Bridge stats course and R: simple tests in R + interpretation.

### Core content

Use same survey dataset.

* T-test (fear by gender):

```{r t-test}
t.test(fear ~ gender, data = small)
```

  * Focus on: group means, t value, df, p-value.

* Chi-square (victim × gender):

```{r chi-square}
tab <- table(small$gender, small$victim)
chisq.test(tab)
```

  * Point out: observed counts, expected counts, p-value.

* Always force interpretive sentences, e.g.:
  "Women reported significantly higher fear than men, *t*(198) = 2.45, *p* = .015."

---

## Session 6 – Correlation & simple regression

### Goals

* Introduce `cor()` and `lm()` for very simple models.

### Core content

* City-level dataset, e.g. `crime_cities` with `unemployment`, `burglary_rate`, `police_per_1000`.
* Correlation:

```{r correlation}
cor(crime_cities$unemployment, crime_cities$burglary_rate, use = "complete.obs")
```

* Scatterplot:

```{r scatterplot}
ggplot(crime_cities, aes(x = unemployment, y = burglary_rate)) +
  geom_point()
```

* Linear regression:

```{r linear-regression}
model1 <- lm(burglary_rate ~ unemployment, data = crime_cities)
summary(model1)
```

* Focus on **interpretation**: direction, strength, practical meaning, not formulas.

---

# Phase 4 – Visualization & APA-style Tables (Sessions 7–8)

## Session 7 – Visualizing criminology data with ggplot2

### Goals

* Equip them to make publication-quality graphs.

### Core content

* Histograms, bar plots, boxplots, scatterplots:

```{r ggplot-examples}
ggplot(small, aes(x = fear)) +
  geom_histogram(binwidth = 1)

ggplot(small, aes(x = gender, y = fear)) +
  geom_boxplot()

ggplot(crime_cities, aes(x = police_per_1000, y = burglary_rate)) +
  geom_point()
```

* Titles & labels:

```{r ggplot-labels}
ggplot(small, aes(x = gender, y = fear)) +
  geom_boxplot() +
  labs(
    title = "Fear of crime by gender",
    x = "Gender",
    y = "Fear of crime (1–5)"
  )
```

### Homework

* Each student produces 2–3 plots + short captions.

---

## Session 8 – APA tables directly from R

### Goals

* Show how to generate APA-style tables from R (no manual Word formatting).

### Suggested packages

* `sjPlot` (simple, friendly), and optionally `apaTables` or `gtsummary`.

```{r install-sjplot}
install.packages("sjPlot")
library(sjPlot)
```

### Examples

* Correlation table:

```{r corr-table}
library(sjPlot)
tab_corr(small %>% select(fear, trust_police, age),
         triangle = "lower",
         digits = 2)
```

* T-test / group means:
  * Use `dplyr` for grouped means (from Session 4) and format with `knitr::kable()`:

```{r means-table}
library(knitr)

table_means <- small %>%
  group_by(gender) %>%
  summarise(
    mean_fear = mean(fear, na.rm = TRUE),
    sd_fear   = sd(fear, na.rm = TRUE),
    n         = n()
  )

kable(table_means, digits = 2, caption = "Fear of crime by gender")
```

* Regression table:

```{r regression-table}
model1 <- lm(fear ~ victim + gender + age, data = small)
tab_model(model1, show.ci = FALSE)
```

These tables will appear nicely in RMarkdown and can be exported to Word/PDF.

---

# Phase 5 – RMarkdown & Reproducible Thesis Writing (Sessions 9–10)

## Session 9 – Introduction to RMarkdown (RMD → Word/PDF)

### Goals

* Get students comfortable with RMarkdown documents and knitting.

### Core content

1. **Create new RMarkdown file**
   * `File → New File → R Markdown…`
   * Choose "Word" or "PDF" as default output.

2. **Explain structure**
   * YAML header:

```yaml
---
title: "Fear of Crime in Ghent"
author: "Student Name"
date: "`r Sys.Date()`"
output: word_document
---
```

   * Text sections using `# Introduction`, `## Methods`, etc.
   * Code chunks:

````markdown
```{r load-packages}
library(tidyverse)
library(sjPlot)
survey <- read_csv("fear_of_crime.csv")
```
````

3. **Inline R results**
   * E.g. in the text:
     "The sample contained `` `r
nrow(survey)` `` respondents."

4. **Knit to Word**
   * Click **Knit** → generates `.docx`.
   * They can share this Word file with supervisor.

### Homework

* Small Rmd report: Intro + Methods + basic Descriptives section with one table + one plot, knitted to Word.

---

## Session 10 – Building an APA-style thesis/paper with RMarkdown

### Goals

* Show them the full pipeline: analysis + APA table + text + references → Word/PDF.

### Core content

1. **Thesis/paper structure in Rmd**

   Use a template you provide with sections:

```markdown
# Introduction

# Methods
## Sample
## Measures
## Procedure

# Results

# Discussion

# References
```

2. **Add analysis in chunks**

````markdown
```{r analysis, echo=FALSE}
model1 <- lm(fear ~ victim + gender + age, data = small)
tab_model(model1, show.ci = FALSE)
```
````

3. **Inline stats in APA style**

   Example in text:

```markdown
Victims reported significantly higher fear of crime than non-victims,
*t*(`r round(model_ttest$parameter, 2)`) = `r round(model_ttest$statistic, 2)`,
*p* = `r signif(model_ttest$p.value, 2)`.
```

   (You can pre-compute `model_ttest <- t.test(fear ~ victim, data = small)` in a chunk.)

4. **References & citations**

   * Provide a `references.bib` file with some example sources.
   * In YAML:

```yaml
---
title: "Fear of Crime in Ghent"
author: "Student Name"
output: word_document
bibliography: references.bib
csl: apa.csl    # APA 7 style file
---
```

   * In text: `@AuthorYear` for citations, e.g.
     "Fear of crime is strongly related to perceived disorder @Skogan1990."

5. **Output to Word or PDF**

   * For **Word**:

```yaml
output: word_document
```

   * For **PDF**:

```yaml
output: pdf_document
```

   * Show both once in class so they see the flexibility.

---

# Phase 6 – Final Mini-Thesis / Paper Project

**Goal:** Students do a **small, reproducible R-based research project** from start to finish.

## What they must deliver

1. **Data file** (given by you or collected in class).
2. **R script** (optional, for data cleaning).
3. **RMarkdown file (.Rmd)** containing:
   * Clear sections (Intro, Methods, Results, Discussion).
   * At least:
     * 2 APA-style tables (e.g., descriptives + regression or t-test).
     * 2 figures (e.g., histogram, boxplot, scatterplot).
   * Inline reporting of key statistics in APA format.
   * At least 5 references in `references.bib`.
4. **Final output**: knitted **Word document** (`.docx`) or **PDF**.

## Example criminology questions

* Is fear of crime higher among women than men?
* How is trust in police related to fear of crime?
* Does victimization predict fear of crime after controlling for age and gender?

---

# Package Checklist (for instructors)

Install once and reuse in all examples:

```{r package-install}
install.packages(c(
  "tidyverse",   # dplyr, ggplot2, readr, etc.
  "sjPlot",      # APA-ish tables for models
  "knitr",       # kable()
  "kableExtra",  # prettier tables (optional)
  "tinytex"      # for PDF if needed
))
```

Load in RMarkdown:

```{r package-load}
library(tidyverse)
library(sjPlot)
library(knitr)
library(kableExtra)
```

---

# Next Steps

If you want, the next steps could be:

* Draft a **template RMarkdown file** for your students (with placeholders for each section + example chunks), and
* Give you **one complete worked example** (tiny criminology study) that you can show in class from raw CSV → final APA-style Word document.

---

# Summary Table: Course Overview

| Session | Topic | Key Skills | Deliverable |
|---------|-------|------------|-------------|
| 0 | Installation & Setup | R/RStudio installation | Working environment |
| 1 | R as a tool | Basic syntax, objects, vectors | Script with calculations |
| 2 | Data frames & datasets | Import data, inspect data | Data loading script |
| 3 | Data wrangling | select(), filter(), mutate() | Subset analysis |
| 4 | Descriptive statistics | Frequency tables, grouped means | Descriptive tables |
| 5 | T-tests & chi-square | Hypothesis testing in R | Statistical tests |
| 6 | Correlation & regression | cor(), lm() | Simple regression model |
| 7 | Data visualization | ggplot2 graphs | Publication-quality plots |
| 8 | APA tables | sjPlot, kable() | Formatted tables |
| 9 | RMarkdown basics | Create Rmd, knit to Word | Simple Rmd report |
| 10 | Full APA paper | Complete workflow | APA-style document |
| Final | Mini-thesis project | Full research pipeline | Complete thesis/paper |

---

*This course plan is designed to be flexible and adaptable to your specific needs and student backgrounds.*
